---
title: "limma_fgsea"
date: '`r format(Sys.Date(), "%m/%d/%Y")`'
output:
  html_document:
    toc: yes
    df_print: paged
geometry: margin=0.25cm
---

```{r setup, include=FALSE}
library(knitr)
knit_hooks$set(crop = knitr::hook_pdfcrop)
opts_chunk$set(echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.align="center", fig.width=8, fig.height=8)
```

```{r}
library(tximport)
library(limma)
library(DESeq2)
library(edgeR)
library(fgsea)
library(biomaRt)
library(tidyverse)
library(gplots)
library(RColorBrewer)
library(DT)
library(reshape)
```

```{r}
#########################################
# Begin: Configuration setting
#########################################
# sample design file name
design_file_name <- "/home/mali/study/Project/CAMSC_Coffman/results/salmon/samples.txt"

# define species being used
species <- "Homo Sapiens"

# get current working directory
dir <- "/home/mali/study/Project/CAMSC_Coffman/results/salmon"
# list files in cwd
list.files(dir)
```

```{r}
# define which biomart object to use
# GRCh37 (Hsapiens.v75) or mouse GRCm38 (Mmusculus.v79))
#define_reference_genome = NULL
mart <- NULL

#define_reference_genome <- "EnsDb.Hsapiens.v75"
mart <- biomaRt::useMart(biomart = "ensembl",
					               dataset = "hsapiens_gene_ensembl",
					               host = "https://grch37.ensembl.org")
```

```{r}
# read in sample design file and conditions for this
# analysis
samples <- read.table(file.path(design_file_name), header = TRUE)
```

# Configuration Info: Study Design File
```{r, results='show'}
display_study_design <- as.data.frame(samples)
colnames(display_study_design)[1] <- "id"
DT::datatable(display_study_design)
```

```{r}
# create list of files to be imported
files <- file.path(dir, samples$mpps_id, "quant.sf")
names(files) <- samples$shortname

# Since there this no EnsDb for Danio Rerio, we are going to building the txi 
# from scratch using biomaRt (By Li Ma 05/18/2023)
####################Start##############################
tx2gene <- getBM(
	attributes = c("ensembl_transcript_id", "ensembl_gene_id"),
	#filters = "ensembl_gene_id",
	mart = mart)
#anns <- anns[match(nm, anns[, 1]), ]
colnames(tx2gene) <- c("tx_id", "gene_id")
####################End################################

txi <- tximport(files, type = "salmon", tx2gene = tx2gene, ignoreTxVersion = TRUE,
                countsFromAbundance = "lengthScaledTPM")
```

# Plot a heatmap for the gene expression data (Top 3000 most variable genes)
## Expression values were log transformed using rlog function from DESeq2
```{r}
gene_exp <- txi$counts
gene_exp <- round(gene_exp)
rld <- vst(gene_exp) %>%
  as.data.frame()
rld$var <- apply(rld, 1, var)
rld <- rld %>%
  arrange(desc(var)) %>%
  head(3000) %>%
  select(-var)
matrix <- as.matrix(t(rld))
heatmap.2(matrix, dendrogram = "row",Colv = FALSE, 
          key = TRUE, keysize = 1, labCol = FALSE, trace = 'none')
```

```{r, results='show'}
y <- DGEList(txi$counts)

# filtering using the desing information
design <- model.matrix(~-1 + condition, samples)
keep <- filterByExpr(y, design = design)
y <- y[keep, ]

# normalize and run voom transformation
y <- calcNormFactors(y)
v <- voom(y, design = design, plot = TRUE)

```

```{r}
# fit the data with limma
fit <- lmFit(v, design)
```

```{r, results='show'}
# comparison between conditionCAMSC and conditionMSC
contr <- makeContrasts(conditionCAMSC - conditionMSC, levels = colnames(coef(fit)))
fit <- contrasts.fit(fit, contrasts = contr)
fit <- eBayes(fit = fit)
res <- topTable(fit, sort.by = "logFC", n = Inf, p = 0.05)
res <- res %>%
  rownames_to_column(var = "ENSEMBL_ID")
datatable(res, 
          caption = htmltools::tags$caption(style = 'caption-side: top; text-align: center; color:black;  font-size:200% ;','Differential Analysis Results'),
          options = list(scrollX = TRUE))
```

# Filter out the genes with logFC greater than 2
```{r, results='show'}
filtered_genes <- res[abs(res$logFC) >= 2, ]
filtered_genes <- filtered_genes
datatable(filtered_genes, 
          caption = htmltools::tags$caption(style = 'caption-side: top; text-align: center; color:black;  font-size:200% ;','Filtered Differential Anlysis Results'),
          options = list(scrollX = TRUE))
```

# Build data frame for performming fgsea analysis using gene symbols + logFC values.
```{r, results='show'}
# convert ensemble gene ids into gene symbols
gene_symbols <- getBM(filters = "ensembl_gene_id",
                      attributes = c("ensembl_gene_id", "external_gene_name"),
                      values = res$ENSEMBL_ID,
                      mart = mart)
colnames(gene_symbols) <- c("ENSEMBL_ID", "SYMBOL")

# inner join gene_symbols with filtered_genes on "ENSEMBL_ID"
res <- inner_join(res, gene_symbols, by = "ENSEMBL_ID")

# build input dataframe for fgsea
input_df_one <- res %>%
  select(SYMBOL, logFC) %>%
  na.omit() %>%
  distinct() %>%
  group_by(SYMBOL) %>%
  summarize(logFC = mean(logFC))
datatable(input_df_one, 
          caption = htmltools::tags$caption(style = 'caption-side: top; text-align: center; color:black;  font-size:200% ;','fgsea Input Dataframe One'), 
          options = list(scrollX = TRUE))
```

# Perform fgsea analysis and show the pathways with adjusted p-values <= 0.05,
## Pathway data used h.all.v2023.1.Hs.symbols.gmt
```{r, results='show'}
ranks_one <- deframe(input_df_one)

# load the pathways into to a names list
pathways_hallmark <- gmtPathways("/home/mali/study/Project/CAMSC_Coffman/reference_data/fgsea/h.all.v2023.1.Hs.symbols.gmt")

# run fgsea algorithm with 10000 permutations
fgsea_res <- fgsea(pathways = pathways_hallmark, 
                   stats = ranks_one,
                   nperm = 10000)

filtered_fgsea_res <- fgsea_res[fgsea_res$padj <= 0.05, ] %>%
  as_tibble() %>%
  arrange(desc(NES)) %>%
  arrange(padj) %>%
  select(-c(nMoreExtreme))
datatable(filtered_fgsea_res,
          caption = htmltools::tags$caption(style = 'caption-side: top; text-align: center; color:black;  font-size:200% ;','Filtered fgsea Output One(Hallmark)'),
          options = list(scrollX = TRUE))
```


# Build data frame for performming fgsea analysis using gene symbols + t statistic values
```{r, results='show'}
# build input dataframe for fgsea
input_df_two <- res %>%
  select(SYMBOL, t) %>%
  na.omit() %>%
  distinct() %>%
  group_by(SYMBOL) %>%
  summarize(t = mean(t))
datatable(input_df_two, 
          caption = htmltools::tags$caption(style = 'caption-side: top; text-align: center; color:black;  font-size:200% ;','fgsea Input Dataframe Two'),
          options = list(scrollX = TRUE))
```

# Perform fgsea analysis and show the pathways with adjusted p-values <= 0.05,
## Pathway data used h.all.v2023.1.Hs.symbols.gmt
```{r, results='show'}
ranks_two <- deframe(input_df_two)

# run fgsea algorithm with 10000 permutations
fgsea_res <- fgsea(pathways = pathways_hallmark, 
                   stats = ranks_one,
                   nperm = 10000)

filtered_fgsea_res <- fgsea_res[fgsea_res$padj <= 0.05, ] %>%
  as_tibble() %>%
  arrange(desc(NES)) %>%
  arrange(padj) %>%
  select(-c(nMoreExtreme))
datatable(filtered_fgsea_res,
          caption = htmltools::tags$caption(style = 'caption-side: top; text-align: center; color:black;  font-size:200% ;','Filtered fgsea Output Two(Hallmark)'),
          options = list(scrollX = TRUE))
```


# Perform fgsea analysis and show the pathways with adjusted p-values <= 0.05,
## Pathway data used c5.all.v2023.1.Hs.symbols.gmt
```{r, results='show'}
# load the pathways into to a names list
pathways_go <- gmtPathways("/home/mali/study/Project/CAMSC_Coffman/reference_data/fgsea/c5.all.v2023.1.Hs.symbols.gmt")

# run fgsea algorithm with 10000 permutations
fgsea_res <- fgsea(pathways = pathways_go, 
                   stats = ranks_one,
                   nperm = 10000)

filtered_fgsea_res <- fgsea_res[fgsea_res$padj <= 0.05, ] %>%
  as_tibble() %>%
  arrange(desc(NES)) %>%
  arrange(padj) %>%
  select(-c(nMoreExtreme))
datatable(filtered_fgsea_res,
          caption = htmltools::tags$caption(style = 'caption-side: top; text-align: center; color:black;  font-size:200% ;','Filtered fgsea Output Three(GO)'),
          options = list(scrollX = TRUE))
```

# Perform fgsea analysis and show the pathways with adjusted p-values <= 0.05,
## Pathway data used c5.all.v2023.1.Hs.symbols.gmt
```{r, results='show'}
ranks_two <- deframe(input_df_two)

# run fgsea algorithm with 10000 permutations
fgsea_res <- fgsea(pathways = pathways_go, 
                   stats = ranks_one,
                   nperm = 10000)

filtered_fgsea_res <- fgsea_res[fgsea_res$padj <= 0.05, ] %>%
  as_tibble() %>%
  arrange(desc(NES)) %>%
  arrange(padj) %>%
  select(-c(nMoreExtreme))
datatable(filtered_fgsea_res,
          caption = htmltools::tags$caption(style = 'caption-side: top; text-align: center; color:black;  font-size:200% ;','Filtered fgsea Output Four(GO)'),
          options = list(scrollX = TRUE))
```

